---
title: "Math 189 final"
output: pdf_document
---


```{r}
cells = read.csv('/Users/Jonnie/Desktop/CERN/train_100_events/event000001000-cells.csv')
hits = read.csv('/Users/Jonnie/Desktop/CERN/train_100_events/event000001000-hits.csv')
particles = read.csv('/Users/Jonnie/Desktop/CERN/train_100_events/event000001000-particles.csv')
truth = read.csv('/Users/Jonnie/Desktop/CERN/train_100_events/event000001000-truth.csv')
```


### Change types
```{r}
for (i in 1:length(cells)){
  print(paste('cells:',names(cells)[i],':',class(cells[,i])))
}
for (i in 1:length(hits)){
  print(paste('hits:', names(hits)[i],':',class(hits[,i])))
}
for (i in 1:length(particles)){
  print(paste('particles:', names(particles)[i],':',class(particles[,i])))
}
for (i in 1:length(truth)){
  print(paste('truth:', names(truth)[i],':',class(truth[,i])))
}
```

```{r}
cells$hit_id = factor(cells$hit_id)
cells$ch0 = as.numeric(cells$ch0)
cells$ch1 = as.numeric(cells$ch1)
hits$hit_id = factor(hits$hit_id)
hits$volume_id = factor(hits$volume_id)
hits$layer_id = factor(hits$layer_id)
hits$module_id = factor(hits$module_id)
particles$particle_id = factor(particles$particle_id)
truth$particle_id = factor(truth$particle_id)


```

### Particles + truth

```{r}
particles_truth <- merge(particles,truth,by="particle_id")
```






```{r}
particles_truth$net_x = abs(particles_truth$vx - particles_truth$tx)
particles_truth$net_y = abs(particles_truth$vy - particles_truth$ty)
particles_truth$net_z = abs(particles_truth$vz - particles_truth$tz)
particles_truth$two_norm = sqrt(particles_truth$net_x^2 + particles_truth$net_y^2 + particles_truth$net_z^2)
```

### orginal and last position

### net displacement
```{r}
df = particles_truth[1,]
df = df[-1,]
i = 1
while (i < nrow(particles_truth)){
  key = particles_truth$particle_id[i]
  test = particles_truth[particles_truth$particle_id == key,]
  df = rbind(df,test[nrow(test),])
  i = i + nrow(test)
  if(nrow(df) > 500){
    break
  }
}
```

```{r}
df$q = factor(df$q)
class(df$q)
```


```{r}
library(ggplot2)
library(latex2exp)
df_plot = qplot(df$two_norm,bins =15 )
df_plot+ labs(title = "Distribution of Total displacement") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$last position - origin positon$||_2$")) 
```




Split positive and negative charge data
```{r}
pos_charge = df[df$q == 1,]
neg_charge = df[df$q ==-1,]
```


```{r}
ggplot() + geom_histogram(data = pos_charge, aes(x = two_norm,color = 'blue'),bins = 15) + geom_histogram(data = neg_charge, aes(x = two_norm,color = 'red'),bins = 15) + labs(title = "Distribution of Total displacement: Positive vs Negatively charged") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$last position - origin positon$||_2$"))
```


Run test 
```{r}
mean(neg_charge$two_norm)
mean(pos_charge$two_norm)
wilcox.test(neg_charge$two_norm, pos_charge$two_norm)
```

Indeed distributions are similar when normalized 
```{r}
ggplot() + geom_histogram(data = pos_charge, aes(x = two_norm,y=..count../sum(..count..),color = 'blue'),bins = 15) + geom_histogram(data = neg_charge, aes(x = two_norm,y=..count../sum(..count..),color = 'red'),bins = 15) + labs(title = "Normalized Distribution: Positive vs Negatively charged") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$last position - origin positon$||_2$"))
```





Run chi-square to see if two_norm data follows expoenential - Fail
```{r}
df_test = layer_data(df_plot,1) 
p2 = pexp(df_test$xmax,rate = 1/mean(df$two_norm)) - pexp(pmax(df_test$xmin,0),rate = 1/mean(df$two_norm))
```

```{r}
chisq.test(df_test$count, p = p2, rescale.p = TRUE)
```

### first displacement
```{r}
df = particles_truth[1,]
df = df[-1,]
i = 1
while (i < nrow(particles_truth)){
  key = particles_truth$particle_id[i]
  test = particles_truth[particles_truth$particle_id == key,]
  df = rbind(df, test[1,])
  i = i + nrow(test)
  if(nrow(df) > 500){
    break
  }
}
```


```{r}
df$q = factor(df$q)
class(df$q)
```


```{r}
library(ggplot2)
df_plot = qplot(df$two_norm,bins =15 )
df_plot + labs(title = "Distribution of first displacement") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$first hit position - origin positon$||_2$"))
```

Split positive and negative charge data
```{r}
pos_charge = df[df$q == 1,]
neg_charge = df[df$q ==-1,]
nrow(pos_charge)
nrow(neg_charge)
```

```{r}
ggplot() + geom_histogram(data = pos_charge, aes(x = two_norm,color = 'blue'),bins = 15) + geom_histogram(data = neg_charge, aes(x = two_norm,color = 'red'),bins = 15)  + labs(title = "Distribution of first displacement: Positive vs Negatively charged") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$first hit position - origin positon$||_2$"))
```



Run test 
```{r}
mean(neg_charge$two_norm)
mean(pos_charge$two_norm)
wilcox.test(neg_charge$two_norm, pos_charge$two_norm)
```
Normalized
```{r}
ggplot() + geom_histogram(data = pos_charge, aes(x = two_norm,y=..count../sum(..count..),color = 'blue'),bins = 15) + geom_histogram(data = neg_charge, aes(x = two_norm,y=..count../sum(..count..),color = 'red'),bins = 15)  + labs(title = "Normalized Distribution: Positive vs Negatively charged") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$first hit position - origin positon$||_2$"))
```


```{r}
df_test = layer_data(df_plot,1) 
p2 = pexp(df_test$xmax,rate = 1/mean(df$two_norm)) - pexp(pmax(df_test$xmin,0),rate = 1/mean(df$two_norm))
```

```{r}
chisq.test(df_test$count, p = p2, rescale.p = TRUE)
```
>
 Same results for First displacement



### Unique values in each event
```{r}
number_particles = c()
for( i in 0:99){
  event = 1000 + i
  event_string = toString(event)
  link = paste0('/Users/Jonnie/Desktop/CERN/train_100_events/event00000',event_string,'-particles.csv')
  particles2 = read.csv(link)
  number_particles = c(number_particles,nrow(particles2))
}
```



```{r}
number_in_event = qplot(number_particles,bins = 10)
number_in_event  + labs(title = "Number of particles across events") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("Number of particles"))
```
### Chi-square
```{r}
event_test = layer_data(number_in_event,1) 
p1 = pnorm(event_test$xmax,mean = mean(number_particles),sd = sd(number_particles)) - pnorm(event_test$xmin,mean = mean(number_particles), sd = sd(number_particles))
```

```{r}
chisq.test(event_test$count, p = p1, rescale.p = TRUE)
```



## KS-test
```{r}
ks.test(number_particles,"pnorm",mean(number_particles),sd(number_particles))
```
 ITS NORMAL


### Particles with no hits
Check numbers of particle ids with no data for truth
```{r}
length(unique(particles$particle_id))
length(unique(particles_truth$particle_id))
length(setdiff(particles$particle_id,truth$particle_id))
```

```{r}
no_truth = particles[particles$particle_id %in% setdiff(particles$particle_id,truth$particle_id),]
nrow(no_truth)
```




### Test proportion of charge in particles with no hits
```{r}
qplot(particles_truth$q,bins = 10)  + labs(title = "Charge of particles for those with hits") + theme(plot.title = element_text(hjust = 0.5)) +xlab("Charge of particle")
qplot(no_truth$q,bins = 10)+ labs(title = "Charge of particles for those with no hits") + theme(plot.title = element_text(hjust = 0.5)) +xlab("Charge of particle")
no_truth_prop = nrow(no_truth[no_truth$q == 1,])/length(no_truth$q)
yes_truth_prop = nrow(particles_truth[particles_truth$q == 1,])/length(particles_truth$q)
paste0('no_truth_prop: ',no_truth_prop)
paste0('yes_truth_prop: ',yes_truth_prop)

```
Z- Test of proportions
```{r}
p_e = nrow(particles[particles$q == 1,])/length(particles$q)
z_observed = (no_truth_prop - yes_truth_prop)/(sqrt((length(particles$q)*p_e*(1-p_e))/(length(no_truth$q)*length(particles_truth$q))))
paste0('Z score: ',z_observed)
paste0('p-value: ',1-pnorm(z_observed))
```

> 
AHA! Proportions are different! There is more positively charged particles (higher proportion of positively charged particles) in the set where there are no hits!



Visualizing starting positions of particles with 0 hits
```{r}
library(plotly)
packageVersion('plotly')

plot_ly(no_truth,x =  ~vx,y = ~vy,z = ~vz,color = ~q,colors =c('blue','red') )
```



### Momentum of particles with no hits
```{r}
no_truth$Mom_norm = sqrt(no_truth$px^2 + no_truth$py^2 + no_truth$pz^2)
particles_truth$Mom_norm = sqrt(particles_truth$px^2 + particles_truth$py^2 + particles_truth$pz^2)
```

```{r}
head(no_truth)
```


```{r}
qplot(no_truth$Mom_norm,bins = 10) + labs(title = "Distribution of momentum for no hits") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$Momentum$||_2$"))
qplot(particles_truth$Mom_norm,bins = 10)  + labs(title = "Distribution of momentum for particles with hits") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$Momentum$||_2$"))
paste0('no_truth: ',mean(no_truth$Mom_norm)) 
paste0('particles_truth: ',mean(particles_truth$Mom_norm)) 
```
> 
Momentum norm of particles seem to be around 0 regardless of having any hits or not


Run test 
```{r}
wilcox.test(no_truth$Mom_norm, particles_truth$Mom_norm)
```



### Momentum predicting nhits


```{r}
mom_pred_nhits = lm(data = particles, nhits~Mom_norm)
summary(mom_pred_nhits)
temp_var <- predict(mom_pred_nhits, interval="prediction")
new_df <- cbind(particles, temp_var)
bob = ggplot(data = new_df , aes(x=Mom_norm, y=nhits)) +
  geom_point() +    
  geom_line(aes(y=lwr), color = "red", linetype = "dashed")+
  geom_line(aes(y=upr), color = "red", linetype = "dashed")+
  geom_smooth(method=lm,se = TRUE)    + labs(title = "Momentum predict nhits") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$Momentum$||_2$"))
bob
```
>
Momentum horrendous at predicting nhits since it clusters at 0 regardless of nhits





Zoom in perhaps?

```{r}
particles_mNorm_less50= particles[particles$Mom_norm < 50,]
mom_pred_nhits = lm(data = particles_mNorm_less50, nhits~Mom_norm)
summary(mom_pred_nhits)
temp_var <- predict(mom_pred_nhits, interval="prediction")
new_df <- cbind(particles_mNorm_less50, temp_var)
bob = ggplot(data = new_df , aes(x=Mom_norm, y=nhits)) +
  geom_point() +    
  geom_line(aes(y=lwr), color = "red", linetype = "dashed")+
  geom_line(aes(y=upr), color = "red", linetype = "dashed")+
  geom_smooth(method=lm,se = TRUE)    + labs(title = "Momentum predict nhits: Zoom") + theme(plot.title = element_text(hjust = 0.5)) +xlab(TeX("$||$Momentum$||_2$"))
bob
```

## Find modes
```{r}
test = hist(particles$nhits,breaks = 20)
mode1 = test$mids[test$counts == max(test$counts)]
mode2 = test$mids[test$counts == sort(test$counts,partial = length(test$counts) - 1)[length(test$counts)-1]]
midpoint = (mode1 + mode2)/2
```
Ignore plot above

```{r}
ggplot(data = particles , aes(x=nhits)) + geom_histogram(bins = 10) + geom_vline(aes(xintercept  = midpoint,color = 'midpoint')) + geom_vline(aes(xintercept = mode1,color = 'mode1')) + geom_vline(aes(xintercept = mode2,color = 'mode2')  ) + labs(title = "Distribution of nhits") + theme(plot.title = element_text(hjust = 0.5)) + scale_color_manual(name = "Labels", values = c(midpoint = "blue", mode1 = 'red', mode2 = "green"))



```

## Chop em 
```{r}
expo_side = particles[particles$nhits < midpoint,]$nhits
norm_side = particles[particles$nhits > midpoint,]$nhits
expo_plot = qplot(expo_side,bins= 7)
norm_plot = qplot(norm_side,bins=7)
expo_plot + labs(title = "Exponential side") + theme(plot.title = element_text(hjust = 0.5))+xlab('nhits')
norm_plot+ labs(title = "Normal side") + theme(plot.title = element_text(hjust = 0.5))+xlab('nhits')
```




```{r}
norm_test = layer_data(norm_plot,1) 
p1 = pnorm(norm_test$xmax,mean = mean(norm_side),sd = sd(norm_side)) - pnorm(norm_test$xmin,mean = mean(norm_side), sd = sd(norm_side))
```

```{r}
chisq.test(norm_test$count, p = p1, rescale.p = TRUE)
```


```{r}
expo_test = layer_data(expo_plot,1) 
p2 = pexp(expo_test$xmax,rate = 1/mean(expo_side)) - pexp(pmax(expo_test$xmin,0),rate = 1/mean(expo_side))
```

```{r}
chisq.test(expo_test$count, p = p2, rescale.p = TRUE)
```

## KS  test
```{r}
ex = rexp(10000, rate = 1/mean(expo_side))
ks.test(ex,"pexp",expo_side)
```

>
FAILURE AGAIN


## Kurtosis Test
```{r}
library('moments')
kurtosis(expo_side)
ex_set = c()
for (i in 1:1000){
  ex = rexp(1000, rate = 1/mean(expo_side))
  ex_k = kurtosis(ex)
  ex_set = c(ex_set,ex_k)
}
qplot(ex_set,bins = 10) + geom_vline(aes(xintercept = kurtosis(expo_side),color = 'sample_value')) + scale_color_manual(name = "Labels", values = c(sample_value = "blue")) + labs(title = "Kurtosis Distribution of exponential with rate of sample") + theme(plot.title = element_text(hjust = 0.5))+xlab('Kurtosis')
```


```{r}
kurtosis(norm_side)
nm_set = c()
for (i in 1:1000){
  nm = rnorm(10000, mean = mean(norm_side), sd = sd(norm_side))
  nm_k = kurtosis(nm)
  nm_set = c(nm_set,nm_k)
}
qplot(nm_set,bins = 10) + geom_vline(aes(xintercept = kurtosis(norm_side),color = 'sample_value'))  + scale_color_manual(name = "Labels", values = c(sample_value = "blue")) + labs(title = "Kurtosis Distribution of Normal with parameters of sample") + theme(plot.title = element_text(hjust = 0.5))+xlab('Kurtosis')
```


### Sum of values by particle_id

```{r}
particles_value <- merge(particles_truth,cells,by="hit_id")
particles_value = aggregate(value ~ hit_id + particle_id+q,data = particles_value,FUN = sum)
particles_value = aggregate(value ~ particle_id+q,data = particles_value,FUN = sum)
```


```{r}
value_plot = qplot(particles_value$value,bins = 10)
value_plot + labs(title = "Distribution of sum of particle values by ID") + theme(plot.title = element_text(hjust = 0.5))+xlab('Particle values')
```

## Chi-square test
```{r}
value_test = layer_data(value_plot,1) 
p2 = pexp(value_test$xmax,rate = 1/mean(particles_value$value)) - pexp(pmax(value_test$xmin,0),rate = 1/mean(particles_value$value))
```

```{r}
chisq.test(value_test$count, p = p2, rescale.p = TRUE)
```


## KS  test
```{r}
ex = rexp(10000, rate = 1/mean(particles_value$value))
ks.test(ex,"pexp",particles_value$value)
```


```{r}
neg_value = particles_value[particles_value$q == -1,]
pos_value = particles_value[particles_value$q ==1,]
```


```{r}
qplot(neg_value$value,bins = 15)+ labs(title = "Distribution for negative charge") + theme(plot.title = element_text(hjust = 0.5))+xlab('Particle value')
qplot(pos_value$value,bins = 15)+ labs(title = "Distribution for positive charge") + theme(plot.title = element_text(hjust = 0.5))+xlab('Particle value')
```


Test if means are diff
```{r}
mean(neg_value$value)
mean(pos_value$value)
wilcox.test(neg_value$value, pos_value$value)
```

```{r}
ggplot() + geom_histogram(data = pos_value, aes(x = value,color = 'blue'),bins = 15) + geom_histogram(data = neg_value, aes(x = value,color = 'red'),bins = 15)+ labs(title = "Positive and negative side by side") + theme(plot.title = element_text(hjust = 0.5))+xlab('Particle value')
```


### sum of all values in all events
```{r}
number_values = c()
for( i in 0:99){
  event = 1000 + i
  event_string = toString(event)
  link = paste0('/Users/Jonnie/Desktop/CERN/train_100_events/event00000',event_string,'-cells.csv')
  cells2 = read.csv(link)
  number_values = c(number_values,sum(cells2$value))
}
```

```{r}
values_plot = qplot(number_values,bins = 15)
values_plot+ labs(title = "Distribution of sum of all values per event") + theme(plot.title = element_text(hjust = 0.5))+xlab('Values per event')
```


### Chi-square
```{r}
values_test = layer_data(values_plot,1) 
p1 = pnorm(values_test$xmax,mean = mean(number_values),sd = sd(number_values)) - pnorm(values_test$xmin,mean = mean(number_values), sd = sd(number_values))
```

```{r}
chisq.test(values_test$count, p = p1, rescale.p = TRUE)
```



## KS-test
```{r}
ks.test(number_values,"pnorm",mean(number_values),sd(number_values))
```



```{r}
ggplot(data = data.frame(values = number_values),aes(sample = values)) + stat_qq() + stat_qq_line(color = "red") + labs(title = "QQ- Norm") + theme(plot.title = element_text(hjust = 0.5)) 
```

